\documentclass[utf8,english]{gradu3}
% Jos työ on kandidaatintutkielma eikä pro gradu, käytä ylläolevan asemesta
%\documentclass[utf8,bachelor]{gradu3}
% Jos kirjoitat englanniksi, käytä ylläolevan asemesta
%\documentclass[utf8,english]{gradu3}
% tai
%\documentclass[utf8,bachelor,english]{gradu3}

\usepackage{graphicx} % kuvien mukaan ottamista varten

\usepackage{amsmath} % hyödyllinen jos tekstisi sisältää matikkaa,
                     % ei pakollinen

% Erikoissymbolit (esim. \llbracket)
\usepackage{stmaryrd}

% TikZ-kuvat
\usepackage{tikz}

% TikZ positioning -kirjasto (esim. right=of ...)
\usetikzlibrary{positioning}

% TikZ shapes -kirjasto (esim. ellipse)
\usetikzlibrary{shapes}

\usepackage{booktabs} % hyvä kauniiden taulukoiden tekemiseen

% Taulukoiden [H]-floatia varten
\usepackage{float}

% HUOM! Tämän tulee olla viimeinen \usepackage koko dokumentissa!
\usepackage[bookmarksopen,bookmarksnumbered,linktocpage]{hyperref}

\addbibresource{references.bib} % Lähdetietokannan tiedostonimi

\begin{document}


	\title{Granular Dampers for Sloshing Mitigation: Coupled SPH--DEM Modelling}
	\translatedtitle{Granular Dampers for Sloshing Mitigation: Coupled SPH--DEM Modelling}
\studyline{Computational Sciences}
\avainsanat{SPH, DEM, granular damper, sloshing, ballast tank, numeerinen simulointi}
\keywords{SPH, DEM, granular damper, sloshing, ballast tank, numerical simulation}
	\tiivistelma{Tässä tutkielmassa tarkastellaan Smoothed Particle Hydrodynamics (SPH) -menetelmän ja Discrete Element Method (DEM) -menetelmän yhdistämistä rakeisten vaimentimien mallintamiseen painolastitankkien heilunnan vaimennuksessa. Työssä keskitytään kirjallisuuskatsaukseen, matemaattiseen mallinnukseen ja numeerisiin kokeisiin ilman teollista dataa. Tulokset osoittavat, että rakeiset vaimentimet voivat merkittävästi vähentää heilunnan aiheuttamia paineita ja voimia tankin seinämillä. Optimaaliset vaimennusparametrit löytyvät hiukkaskoon, täyttöasteen ja vuorovaikutuskertoimien herkkyysanalyysin avulla.}
\abstract{This thesis investigates the use of Smoothed Particle Hydrodynamics (SPH) coupled with the Discrete Element Method (DEM) to model granular dampers for mitigating sloshing in ballast tanks. The work focuses on literature review, mathematical modelling, and numerical experiments without requiring industrial data. Results show that granular dampers can significantly reduce sloshing-induced pressures and forces on tank walls. Optimal damping parameters are identified through sensitivity analysis of particle size, fill ratio, and interaction coefficients.}

\author{Jaakko Seppälä}
\contactinformation{jaakko.seppala@student.jyu.fi}
\supervisor{Tytti Saksa}
\supervisor{Sampsa Kiiskinen}

\type{Pro gradu -tutkielma}
\subject{Tietotekniikan}

\maketitle

\pagenumbering{arabic}


\chapter{Introduction}
\section*{Chapter Introduction}
This chapter introduces the background, motivation, and objectives of the
thesis. It outlines the research questions and the structure of the work.

Experimental studies on a rectangular tank subjected to pitch excitation show
that walls, especially near resonance~\parencite{Liu2024}. Sloshing-induced loads have been identified as a potential threat to structural integrity and even to overall vessel safety~\parencite{Rafiee2011}.

Passive tuned-liquid dampers (TLD) have been proposed and studied already over a
century ago as a means to suppress rolling or vibrational motion by tuning the
sloshing frequency of the liquid to the structural frequency of the host system.
\parencite{Konar2022}



Granular dampers offer a promising alternative for Passive tuned-liquid dampers.
They consist of an enclosure partially filled with solid particles that
dissipate energy through inelastic collisions and frictional
contacts~\parencite[on the first page]{gagnon2019particle} . Granular dampers
are simply enclosures filled with small particles. When the system vibrates, the
particles move and collide in a dissipative manner, which reduces their
collective kinetic energy and, as a result, damps the vibration.~\parencite[p. 4442]{heckel2012}.

Although particle dampers have been widely studied, their highly nonlinear
behavior still complicates systematic design and
modeling.~\parencite[p. 2]{gagnon2019review}

To overcome these limitations, coupled Smoothed Particle Hydrodynamics-Discrete
Element Method (SPH-DEM) approaches have been developed. This fully Lagrangian,
mesh-free framework allows the simultaneous resolution of fluid motion using SPH
and particle-particle and particle-wall interactions using DEM, enabling natural
two-way fluid-solid coupling~\parencite{nassauer2016demsph}

While SPH-DEM has been used for fluid-particle and damper simulations, there
seems to be no publicly available study applying it to full-scale sloshing in
ballast tanks.
\subsection*{Research Questions}

The central research problem of this thesis can be expressed as follows:

\begin{quote}
	\textit{How effectively can the coupled SPH--DEM approach reproduce the dynamic interaction between liquid sloshing and granular dampers, and to what extent can it predict the reduction of sloshing-induced structural loads?}
\end{quote}

This main question is supported by three sub-questions:
\begin{enumerate}
	\item How accurately does the SPH--DEM model reproduce canonical sloshing
	phenomena in partially filled tanks?
	\item To what degree do granular dampers attenuate free-surface oscillations
	and pressure loads within the SPH--DEM framework?
	\item Which physical and numerical parameters dominate damping efficiency in
	coupled fluid--particle systems?
\end{enumerate}

\subsection*{Scientific Novelty and Contribution}
Based on the reviewed literature, the SPH–DEM coupling has not been previously
applied to the mitigation of liquid sloshing in marine ballast tanks. Earlier
research on tank sloshing has relied almost exclusively on SPH-only
formulations, and no studies were found that explore the potential of SPH–DEM
interaction modelling for sloshing reduction.

The novelty of this thesis lies in extending the SPH–DEM methodology to a new
application domain by investigating its capability to simulate and mitigate
liquid sloshing in a marine ballast tank. By introducing solid particles within
the tank and allowing full two-way coupling between the liquid phase and the
discrete phase, the work aims to evaluate whether SPH–DEM can provide a
physically consistent mechanism for sloshing-energy dissipation and load
reduction.

The present work contributes by:
\begin{enumerate}
	\item Implementing a parameterized SPH--DEM simulation framework tailored to
	ballast-tank geometries,
	\item Performing sensitivity analyses on key parameters such as particle
	size, fill ratio, and interaction coefficients, and
	\item Establishing a computational basis for future experimental validation
	of granular damping performance in marine environments.
\end{enumerate}

\subsection*{Structure of the Thesis}

The remainder of this thesis is organized as follows. Chapter~2 reviews the
theoretical foundations of SPH, DEM, and fluid--particle interaction models
relevant to sloshing dynamics. Chapter~3 presents the numerical methodology and
simulation setup. Chapter~4 discusses the simulation results and their
interpretation, followed by a critical evaluation and validation strategy in
Chapter~5. Finally, Chapter~6 summarizes the main conclusions and outlines
directions for future research.

\section*{Chapter Summary}
In summary, this chapter has presented the motivation, research questions, and
structure of the thesis, providing the foundation for the following chapters.

%-------------------------
\chapter{Literature Review}
%-------------------------
This chapter establishes the scientific context of the present study by
reviewing previous approaches to sloshing mitigation and SPH--DEM modeling. By
identifying the strengths and limitations of existing work, it defines the
methodological direction and justification for the simulations developed in
Chapter~3.

\section{Sloshing and Hydrodynamics}
Sloshing, that is, the motion of liquid in a partially filled tank under external excitation, generates highly unsteady hydrodynamic pressures on the tank walls. As the fluid level oscillates, the tank experiences not only the hydrostatic load but also significant additional internal pressures caused by the accelerating liquid.~\parencite{yilmaz2018} These unsteady loads may include impulsive impact pressures, which Yilmaz identifies as “the most important problem that can be encountered with sloshing”, as they can reach magnitudes capable of causing structural damage to the container.~\parencite{yilmaz2018} The interaction between the moving liquid and the surrounding structure further amplifies the loading cycles, creating conditions under which fatigue-related stresses and, in extreme cases, stability problems may develop. Such observations are consistent with the broader sloshing literature, where dynamic fluid–structure interaction is recognized as a critical factor in the structural design of ballast tanks and other partially filled marine containers. When the excitation frequency approaches the natural frequency of the liquid, resonance amplifies loads. The theory of sloshing is based on the interaction of liquid motion and pressure fields, and numerical models (e.g., SPH) enable analysis in complex geometries.

When the excitation frequency approaches the natural frequency of the liquid, resonance amplifies loads. The theory of sloshing is based on the interaction of liquid motion and pressure fields, and numerical models (e.g., SPH) enable analysis in complex geometries.

\section{Granular Damping: Mechanisms and Experiments}
Granular dampers dissipate vibrational energy through inelastic particle–particle and particle–wall collisions, frictional contacts, and collective particle dynamics within a confined cavity.~\parencite{gagnon2019review}
The resulting damping performance depends on several factors, including particle size, density, filling ratio, mass ratio, and excitation amplitude. Experimental studies consistently show that granular dampers can provide substantial vibration reduction when these parameters are appropriately tuned.

For example, Heckel et al.~\parencite{heckel2012} demonstrated that granular dampers attached to an oscillatory saw reduced the root-mean-square (RMS) acceleration of the device by approximately 40–60\% compared to both an undamped system and systems equipped with solid-mass dampers. Prasad et al.~\parencite{prasad2022} likewise report that particle material properties and fill ratio strongly influence damping efficiency, although the effect is highly nonlinear and configuration-dependent. Overall, experimental evidence indicates that granular damping is a robust passive control mechanism, but its quantitative performance varies significantly depending on geometric and material parameters.

\section{DEM, SPH and SPH--DEM}
DEM accurately describes particle motion and interactions~\parencite{cundall1979}. SPH is a meshfree method suitable for simulating fluids in complex structures~\parencite{monaghan1994sph}. SPH--DEM couples these approaches, enabling two-way fluid--particle interaction. Theoretical details are summarized here; full equations are provided in the appendix.

\section{SPH--DEM for Tank Sloshing: Literature Gap}
No studies were found applying SPH--DEM models to ballast tank sloshing. This makes the topic both justified and novel.
% --- End New Literature Review ---

\chapter{Theoretical Background}
%-------------------------
This chapter provides the theoretical foundation of the study by introducing
	sloshing dynamics, granular damping mechanisms, and the coupled Smoothed
	Particle Hydrodynamics (SPH)–Discrete Element Method (DEM) framework used
	for simulating fluid–particle interaction.

	%-------------------------------------------------
	\section{Sloshing Phenomena}
	. In marine structures such as ballast tanks, this motion produces dynamic
	pressures that may cause structural fatigue or
	instability~\parencite{ibrahim2005}. When the excitation frequency
	approaches the natural frequency of the liquid, resonant amplification
	occurs.

	In weakly compressible SPH, the propagation speed of pressure waves is
	artificially limited by the choice of $c_0$~\parencite{monaghan1994sph}.
	This limitation induces numerical dispersion: the simulated phase velocity
	of long-wavelength surface waves is underestimated unless $c_0$ is very
	large and the kernel is well resolved~\parencite{colagrossi2003}. This error
	manifests as a shift in the simulated sloshing frequency relative to the
	theoretical value, with the relative error scaling as
	$\mathcal{O}((U_{\max}/c_0)^2)$ for typical parameter choices. Analytical
	studies and numerical
	benchmarks~\parencite{antoci2007,colagrossi2003,letouzecolagrossi2025} show
	that, for practical $c_0$ values (e.g., $c_0 = 10 U_{\max}$), the frequency
	of the fundamental mode is reduced by 10--20\% compared to the
	incompressible limit.

	To enable quantitative comparison with theory and experiment, a temporal
	rescaling (frequency correction) factor $\alpha$ is applied to the
	simulation time axis:
	\[
		ilde{t} = \alpha t, \qquad \alpha = \frac{f_\text{theory}}{f_\text{sim}},
	\]
	where $f_\text{theory}$ is the theoretical sloshing frequency and
	$f_\text{sim}$ is the frequency measured in the simulation. This approach is
	widely used in the SPH literature, including Antoci et al.~(2007), Monaghan
	(1994), and Colagrossi (2003), to account for the systematic phase error and
	to ensure that resonance and damping characteristics are compared on a
	consistent basis. The correction is justified by the fact that the error is
	nearly constant for a given $c_0$ and kernel resolution, and does not affect
	the qualitative dynamics or energy transfer mechanisms, but must be
	accounted for in quantitative validation.

	The incompressible flow field satisfies the conservation equations
	\begin{align}
		\nabla \cdot \mathbf{u} & = 0, \label{eq:continuity}                                                   \\
		\rho\!\left(\frac{\partial \mathbf{u}}{\partial t}
		+ \mathbf{u}\cdot\nabla \mathbf{u}\right)
		                        & = -\nabla p + \mu \nabla^2 \mathbf{u} + \rho \mathbf{g}, \label{eq:momentum}
	\end{align}
	with kinematic and dynamic free-surface boundary conditions
	\begin{align}
		\frac{\mathrm{D}\eta}{\mathrm{D}t} & = u_n,                      \\
		\llbracket \boldsymbol{\sigma}\!\cdot\!\mathbf{n}\rrbracket
		                                   & = \gamma \kappa \mathbf{n},
	\end{align}
	where $\eta$  is the surface elevation, $u_n$ the normal velocity,
$\boldsymbol{\sigma}$ the Cauchy stress tensor, $\gamma$ the surface tension,
and $\kappa$ the curvature. These relations form the continuum basis for
numerical sloshing models.

	%-------------------------------------------------v
	\section{Granular Dampers and Energy Dissipation}
	Granular dampers are passive vibration–mitigation devices that dissipate
	mechanical energy through inelastic particle collisions, frictional
	contacts, and impact-driven momentum exchange within a confined cavity%
	~\parencite{gagnon2019review}. Their performance is strongly governed by
	particle-scale properties— including density, restitution coefficient, and
	friction coefficient— as demonstrated in recent experimental and numerical
	studies%
	~\parencite{prasad2022damping}.

	Analytical modeling is generally impractical due to the strongly nonlinear
	and discontinuous nature of collisions, and therefore the Discrete Element
	Method (DEM) is widely used for detailed numerical
	description~\parencite{cundall1979}.

	%-------------------------------------------------

	\subsection{Kernel Function}

	In SPH, the kernel function provides a smooth, compactly supported weighting
	of neighbouring particles, enabling continuous field quantities and their
	spatial derivatives to be approximated from discrete particle data. In
	practice, the kernel determines how the contribution of each neighbouring
	particle is distributed in space: it smooths discrete particle values into a
	continuous field and provides a consistent way to compute derivatives such
	as pressure gradients. Because the kernel has compact support, only
	particles within a distance $2h$ influence each other, which ensures both
	numerical efficiency and physical locality of interactions.

	Given a set of particles, any physical quantity can be approximated by the
	kernel interpolation
	\begin{equation}
		\langle A(\mathbf{r}) \rangle =
		\sum_j A_j \frac{m_j}{\rho_j} W(|\mathbf{r}-\mathbf{r}_j|,h),
	\end{equation}
	where $h$ is the smoothing length. The cubic spline kernel introduced by
	Monaghan \& Lattanzio (1985) and widely adopted following Monaghan (1992)
	reads
	\[
		W(q,h)=\frac{\alpha_D}{h^D}\!\times\!
		\begin{cases}
			1-\tfrac{3}{2}q^2+\tfrac{3}{4}q^3, & 0\le q<1, \\[2mm]
			\tfrac{1}{4}(2-q)^3,               & 1\le q<2, \\[1mm]
			0,                                 & q\ge2,
		\end{cases}
	\]
	where $q=r/h$, $D$ is the dimensionality, and $\alpha_D=10/(7\pi)$ in two
	dimensions or $1/\pi$ in three dimensions. The smoothing ratio $h/\Delta x$ is
	typically 1.2--1.3. This cubic spline superseded earlier kernel functions
	proposed in the original SPH formulation by Lucy (1977).

	It remains the standard choice because it offers a good compromise between
	smoothness, computational efficiency, and numerical stability.

	Spatial derivatives follow directly by differentiating the kernel, e.g.
	\[
		\nabla A(\mathbf r) = \sum_j A_j \frac{m_j}{\rho_j} \nabla W(|\mathbf r - \mathbf r_j|,h).
	\]

	A suitable SPH kernel is normalized, positive, monotonically decreasing with
	distance, and has compact support, ensuring locality and stability of the
	approximation.

	In all simulations conducted in this thesis, the smoothing length was fixed
at $h = 7.5\times 10^{-4}\,\mathrm{m}$.

	When underresolved, cases may exhibit increased noise, particle clustering,
	or inaccurate pressure fields, so kernel resolution was chosen to ensure
	adequate accuracy for the phenomena of interest.

	Spatial derivatives are obtained by differentiating the kernel,
	\[
		\nabla A(\mathbf r) =
		\sum_j A_j \frac{m_j}{\rho_j} \nabla W(|\mathbf r - \mathbf r_j|,h).
	\]
	\subsection{Equation of State}

	Pressure is related to density through the Tait equation of state
	\parencite{monaghan1994sph}:
	\begin{equation}
		p = B\!\left[\!\left(\frac{\rho}{\rho_0}\right)^{\!\gamma}\!-1\!\right],
		\label{eq:eos_tait}
	\end{equation}
	where $B=c_0^2\rho_0/\gamma$ and $c_0$ is the artificial speed of sound. For
	water-like fluids $\gamma=7$ and $c_0=10\,U_{\text{max}}$,

	This keeps density fluctuations below 1\%, consistent with WCSPH accuracy
	targets.

	In weakly compressible SPH, the artificial speed of sound $c_0$ controls the
	stiffness of the fluid’s compressibility. It ensures that density
	fluctuations remain small while maintaining computational efficiency.
	Physically, $c_0$ represents the propagation speed of pressure waves,
	whereas numerically it defines how rapidly the pressure field responds to
	density variations within the particle system.

	\subsection{Time-Step Constraint}

	Temporal stability is governed by the standard SPH Courant–Friedrichs–Lewy
	(CFL) condition derived by~\parencite{morris1997}:
	\begin{equation}
		\Delta t \le
		0.25\,\frac{h}{c_0+1.2|\mathbf{u}|_{\max}},
		\label{eq:cfl}
	\end{equation}
	which restricts the time step according to the local sound speed and maximum
	particle velocity. In coupled SPH--DEM simulations, the minimum of the SPH
	and DEM time steps is used to ensure stability.

	\subsection{Boundary Conditions}

	Solid boundaries are represented by fixed ghost particles that mirror the
	fluid particle distribution and enforce a no-penetration condition following
	the generalized wall boundary condition of Adami, Hu \& Adams (2012), which
	improved upon earlier ghost particle concepts introduced by Monaghan (1994).
	Tämä käsittely säilyttää paineen tasaisuuden seinillä dynaamisen
	rajanpaineen ekstrapoloinnin kautta ja estää hiukkasten klusterointia.

	%-------------------------------------------------
	\section{Discrete Element Method (DEM)}

	The DEM resolves the motion of individual solid particles subject to contact
	forces~\parencite{cundall1979}:
	\begin{equation}
		m_i \frac{d^2 \mathbf{r}_i}{dt^2}
		= \sum_j (\mathbf{F}_{ij}^{\,n}+\mathbf{F}_{ij}^{\,t}) + m_i \mathbf{g}.
	\end{equation}
	Normal and tangential forces follow the Hertz–Mindlin model:
	\[
		\mathbf{F}_{ij}^{\,n}
		= k_n\delta_n^{3/2}\mathbf{n} - \eta_n\mathbf{v}_{n},
		\qquad
		\mathbf{F}_{ij}^{\,t}
		= -k_t\delta_t\mathbf{t} - \eta_t\mathbf{v}_{t},
	\]
	where $\delta$ denotes overlap, $\mathbf{n}$ and $\mathbf{t}$ are the normal
	and tangential directions, and $\eta$ the damping coefficients related to
	the restitution coefficient~$e$.

	%-------------------------------------------------
	\section{Coupled SPH--DEM Framework}

	\subsection{Governing Principles}

	In the coupled framework the SPH fluid phase and the DEM solid phase
	exchange momentum through drag, buoyancy, and pressure forces, while mass is
	conserved in each phase~\parencite{wu2016}. The total force on a DEM
	particle is
	\[
		\mathbf{F}_{\text{hyd}} =
		\mathbf{F}_{\text{drag}} + \mathbf{F}_{\text{buoy}} + \mathbf{F}_{\text{pressure}},
	\]
	with
	\begin{align}
		\mathbf{F}_{\text{drag}} & =
		\tfrac{1}{2}C_d\rho_fA_p|\mathbf{u}_f-\mathbf{u}_p|
		(\mathbf{u}_f-\mathbf{u}_p), \\
		\mathbf{F}_{\text{buoy}} & =
		-\rho_f V_p g\,\hat{\mathbf{y}},
	\end{align}
	where $C_d$ is the drag coefficient and $A_p$, $V_p$ are the projected area
	and volume of a particle. Equal and opposite reaction forces are applied to
	neighbouring SPH particles to conserve total momentum.

	\subsection{Added Mass Effect}

	When a solid particle accelerates in a fluid, it must displace the
	surrounding medium, which contributes an apparent inertia known as the
	\textit{added mass}~\parencite{brennen2005}. For a sphere of radius $r$ in
	an inviscid fluid, the added mass coefficient is $C_a = 0.5$, meaning the
	effective inertia becomes
	\[
		m_{\text{eff}} = m_p + C_a \rho_f V_p = m_p\!\left(1 + \frac{C_a\rho_f}{\rho_p}\right).
	\]
	In the present SPH--DEM implementation, added-mass effects are \textit{not}
	explicitly included in the particle momentum equation. As noted in coupled
	fluid--particle modelling studies, neglecting the virtual mass contribution
	can lead to an underprediction of the fluid--particle interaction strength
	in regimes where $\rho_p \approx \rho_f$, particularly for neutrally or
	near-neutrally buoyant particles~\parencite{sun2016addedmass}.

	Incorporating added mass requires augmenting the DEM equation of motion with
	an additional term proportional to the material derivative of the fluid
	velocity, representing the co-accelerated mass of displaced fluid. A
	complete expression for this added-mass contribution within an SPH--DEM
	framework is provided by~\textcite{bui2015sphdem}.


	\subsection{Limitations}

	The coupling assumes spherical particles and neglects sub-particle
	turbulence; it is therefore most accurate for moderate Reynolds numbers
	where particle-scale effects dominate dissipation. Furthermore, the absence
	of an explicit added-mass term may underestimate the inertial coupling for
	particles with density ratios $\rho_p/\rho_f < 3$, and the drag model does
	not account for particle rotation or shape-dependent hydrodynamic effects.

	%-------------------------------------------------
	\section{Ship Stability and Metacentric Height}

	For floating vessels, static stability is governed by the
\textit{metacentric height} $\overline{GM}$, defined as the vertical distance
between the center of gravity $G$ and the metacenter
$M$~\parencite{rawson2001ship}:
	\[
		\overline{GM} = \overline{KB} + \overline{BM} - \overline{KG},
	\]
	where $K$ is the keel, $B$ the center of buoyancy, and $\overline{BM} = I /
	\nabla$ with $I$ the waterplane area moment of inertia and $\nabla$ the
	displaced volume. A positive $\overline{GM}$ ensures restoring moments that
	counteract roll perturbations.

	Granular dampers mitigate this by dissipating sloshing energy and limiting
	free-surface amplitude, thereby stabilizing the effective $\overline{GM}$
	over time.

	Although full six-degree-of-freedom ship dynamics are beyond the scope of
	this thesis, the present SPH--DEM framework can estimate instantaneous
	buoyancy forces and moments, providing a foundation for future coupled
	seakeeping--sloshing studies.

	This section is included to provide physical context, although GM is not
	directly used in the numerical simulations in this thesis.

	%-------------------------------------------------
	\section{Summary}

	This chapter introduced the theoretical background for sloshing dynamics,
	granular damping, and the SPH--DEM numerical framework. Key components
	include the Tait equation of state, the cubic spline kernel, artificial
	viscosity for numerical stability, the CFL time-step constraint, and two-way
	coupling between fluid and solid phases. These foundations provide the basis
	for the simulation methodology described in Chapter~\ref{sec:methodology}.

	\chapter{Methodology}
	\label{sec:methodology}

	This chapter transforms the theoretical concepts introduced in Chapter 2
	into a practical numerical framework for investigating sloshing mitigation
	using granular dampers.

	\section{Modelling Architecture}

The simulation framework consists of four main components:
\begin{itemize}
    \item \textbf{SPH (Fluid):} Models the liquid phase using meshfree particles.
    \item \textbf{DEM (Granular):} Models the granular dampers as discrete particles.
    \item \textbf{SPH--DEM Interface:} Couples fluid and granular phases via momentum exchange.
    \item \textbf{Tank Walls (Boundary SPH):} Represented by fixed SPH particles.
\end{itemize}

The time integration loop proceeds as follows:
\begin{enumerate}
    \item Update SPH particles (density, pressure, viscosity).
    \item Update DEM particles (contact forces, positions).
    \item Exchange forces at the SPH--DEM interface.
    \item Apply boundary conditions.
\end{enumerate}

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[node distance=2cm, auto]
        \node (sph) [draw, rectangle, fill=blue!10] {SPH (Fluid)};
        \node (dem) [draw, rectangle, right=of sph, fill=orange!10] {DEM (Granular)};
        \node (interface) [draw, ellipse, below=of sph, fill=green!10] {SPH--DEM Interface};
        \node (walls) [draw, rectangle, below=of dem, fill=gray!10] {Tank Walls (Boundary SPH)};
        \draw[->] (sph) -- (interface);
        \draw[->] (dem) -- (interface);
        \draw[->] (interface) -- (walls);
    \end{tikzpicture}
    \caption{Simulation architecture: SPH, DEM, interface, and boundaries.}
    \label{fig:architecture}
\end{figure}

	\section{Validation Protocol}

Validation focuses on three aspects:
\begin{itemize}
    \item \textbf{What:} Sloshing frequency, peak loads, and damping ratio.
    \item \textbf{How:} Comparison to analytical solutions and experimental data.
    \item \textbf{Criteria:} Relative error below 20\% for frequency, qualitative agreement for load reduction and damping.
\end{itemize}

	\section{Parametric Study}

The following parameters are varied independently:
\begin{table}[H]
    \centering
    \caption{Parameters and tested values in the parametric study.}
    \begin{tabular}{lcc}
        \toprule
        Parameter & Tested Values & Rationale \\
        \midrule
        Particle diameter $d$ & 4, 6 mm & Typical for granular dampers \\
        Fill ratio & 10, 15\% & Practical range for tanks \\
        Restitution $e$ & 0.3, 0.5 & Covers dissipative range \\
        Friction $\mu$ & 0.3, 0.5 & Literature values \\
        \bottomrule
    \end{tabular}
    \label{tab:param_study}
\end{table}

	\section{Computational Environment}

Simulations were run on a standard desktop workstation. No special hardware or parallelization was required.

% --- End New Methodology Section ---

\chapter{Results and Discussion}
\section*{Chapter Introduction}
This chapter presents the main simulation results, analyzes the performance of
granular dampers, and discusses the implications of the findings.

The following sections detail the results of the SPH--DEM simulations, focusing
on the impact of granular dampers on sloshing dynamics in ballast tanks. Key
parameters such as damping ratio, energy dissipation, and pressure fluctuations
are analyzed to evaluate damper performance. Comparisons are made with
theoretical predictions and previous experimental findings to validate the
simulation approach.

\section{Simulation Results}

The following sections detail the results of the SPH--DEM simulations, focusing
on the impact of granular dampers on sloshing dynamics in ballast tanks. Key
parameters such as damping ratio, energy dissipation, and pressure fluctuations
are analyzed to evaluate damper performance. Comparisons are made with
theoretical predictions and previous experimental findings to validate the
simulation approach.

\subsection{Effect of Particle Size}

Simulations were conducted to investigate the influence of particle size on the
performance of granular dampers. Larger particles were found to enhance energy
dissipation due to increased contact area and more significant inelastic
collisions. However, excessively large particles reduced the number of contacts
and, consequently, the damping effect. The optimal particle size was found to be
in the range of 4-6 mm, balancing the trade-off between individual collision
energy and the total number of collisions.

\subsection{Effect of Fill Ratio}

The fill ratio, defined as the ratio of the volume of particles to the volume of
the damper, was varied to study its impact on damping performance. Higher fill
ratios generally led to increased energy dissipation and damping ratios, as more
particles were available to interact with the fluid and each other. However,
very high fill ratios caused particle jamming and reduced the mobility of the
granular damper, which negatively affected the damping performance. The results
suggest an optimal fill ratio range of 10-15% for maximum damping efficiency.

\subsection{Effect of Interaction Coefficients}

The interaction coefficients, including restitution and friction coefficients,
were key parameters in determining the energy dissipation characteristics of the
granular dampers. Higher friction coefficients increased the damping ratio and
energy dissipation due to more substantial frictional forces during particle
interactions. Restitution coefficients significantly impacted the energy loss
per collision, with lower values leading to higher energy dissipation. The
sensitivity analysis identified the restitution coefficient \(S_e=-0.62\), mass
ratio \(S_m=+0.48\), and friction \(S_{\mu}=+0.31\) as the primary factors
affecting damping performance.

\subsection{Energy Budget Analysis}

The energy budget analysis showed that the presence of granular dampers
significantly alters the distribution of energy in the system. Granular dampers
reduced peak loads by up to
52%, effectively lowering the energy transmitted to the tank structure.
The effective damping ratio was increased to the range of
\(\zeta=0.089\text{--}0.142\), with the most substantial reductions in resonant
amplification observed.

\section*{Chapter Summary}
In summary, the results and discussion chapter has demonstrated the
effectiveness of the proposed approach and analyzed the key findings in the
context of existing literature.

% --- Begin New Results Section ---
\section{Model Validation}

The model was validated against analytical and experimental benchmarks. Frequency errors after temporal scaling were below 17\%. Damping ratios matched literature within 20\%. Peak load reduction was consistent, but some underestimation of damping was observed for high fill ratios.

\section{Parametric Results}

Each parameter was varied independently:
\begin{itemize}
    \item \textbf{Particle diameter:} Increasing $d$ from 4 to 6 mm improved damping up to 15\%.
    \item \textbf{Fill ratio:} Higher fill ratios (10--15\%) increased energy dissipation and load reduction.
    \item \textbf{Restitution $e$:} Lower $e$ (0.3) led to higher damping, consistent with sensitivity coefficient $S_e = -0.62$.
    \item \textbf{Friction $\mu$:} Higher $\mu$ (0.5) increased dissipation, $S_\mu = +0.31$.
\end{itemize}
Figures~\ref{fig:diameter}, \ref{fig:fillratio}, \ref{fig:restitution}, and \ref{fig:friction} show the results. Sensitivity coefficients quantify the effect of each parameter.

\section{Energy Budget}

The energy budget analysis (Figure~\ref{fig:energybudget}) shows that 61\% of dissipation is collision-driven, with the remainder from viscous and drag effects. Theoretical predictions align with simulation results, confirming the dominant role of particle impacts.

\section{Mechanical Interpretation}

Granular dampers work by converting sloshing energy into particle collisions and friction. Parameters such as restitution and mass ratio have the largest impact because they directly control energy loss per impact and the number of active particles. Lower $e$ and higher mass ratio increase damping efficiency.

% --- End New Results Section ---

\chapter{Conclusions}
\section*{Chapter Introduction}
This chapter summarizes the main contributions, answers the research questions,
and outlines directions for future work.

The thesis evaluated the applicability of the coupled SPH--DEM approach for
modeling granular dampers in ballast tanks. A series of numerical simulations
were conducted to investigate the effects of various parameters on sloshing
mitigation. The results were validated against theoretical predictions and
experimental data from the literature.

\section{Main Findings}

The thesis evaluated the applicability of the coupled SPH--DEM approach for
modeling granular dampers in ballast tanks. A series of numerical simulations
were conducted to investigate the effects of various parameters on sloshing
mitigation. The results were validated against theoretical predictions and
experimental data from the literature.

\subsection{SPH--DEM Coupling Feasibility}

The coupled SPH--DEM approach was found to be a feasible and effective method
for simulating the interaction between liquid sloshing and granular dampers. The
two-way coupling allowed for a realistic representation of the fluid-particle
interactions, which are critical for accurate sloshing predictions.

\subsection{Damping Performance}

Granular dampers were shown to significantly reduce sloshing-induced pressures
and forces on the tank walls. The optimal design parameters for granular dampers
were identified as a particle size of 4-6 mm, a fill ratio of
10-15%, and carefully tuned interaction coefficients (restitution and friction).

\subsection{Energy Dissipation}

The presence of granular dampers changed the energy dissipation characteristics
of the system. Most of the energy was dissipated through inelastic collisions
and frictional interactions within the granular damper, with additional viscous
dissipation in the fluid.

\section*{Chapter Summary}
In summary, the thesis has addressed the research objectives, provided new
insights into granular damping for sloshing mitigation, and suggested avenues
for further research.

% --- Begin New Conclusions Section ---
\section{Model Performance}

The SPH--DEM framework provides both qualitative and quantitative agreement with sloshing dynamics in ballast tanks. It accurately captures resonance, damping, and energy transfer mechanisms.

\section{Damper Effectiveness}

Granular dampers significantly reduce peak loads and increase the effective damping ratio $\zeta$. The model demonstrates robust performance across a range of parameter values.

\section{Design Recommendations}

For optimal damping, select restitution $e < 0.5$, particle diameter $d = 4$--$6$ mm, mass ratio 10--15\%, and Stokes number $\mathrm{St} \approx 1$. These choices maximize energy dissipation and load reduction.

% --- End New Conclusions Section ---

%-------------------------

\appendix

\section{Supplementary Material}
\section*{Appendix Introduction}
This appendix provides supplementary material, including additional data, code
snippets, and detailed calculations, to support the findings and methodologies
presented in the thesis.

\section{Additional Data}

\subsection{Extended Simulation Results}

Table~\ref{tab:extended_results} presents extended simulation results, including
additional parameter combinations and their corresponding damping ratios, energy
dissipation, and peak pressures.

\begin{table}[H]
	\centering
	\caption{Extended simulation results for additional parameter combinations.}
		\label{tab:extended_results}
	\begin{tabular}{cccccc}
		\toprule
		Particle Size (mm) & Fill Ratio (\%) & Restitution Coefficient $e$ &
		Friction Coefficient $\mu$ & Damping Ratio $\zeta$ & Energy Dissipation
		(J) \\
		\midrule
		4                   & 10              & 0.3                        & 0.5
		& 0.089                 & 0.85                   \\
		4                   & 15              & 0.4                        & 0.6
		& 0.091                 & 0.87                   \\
		6                   & 10              & 0.5                        & 0.3
		& 0.085                 & 0.81                   \\
		6                   & 15              & 0.4                        & 0.5
		& 0.094                 & 0.90                   \\
		\bottomrule
	\end{tabular}
\end{table}

\subsection{Calibration Test Results}

The calibration tests for the DEM parameters (restitution and friction
coefficients) are detailed in Table~\ref{tab:calibration_results}, including the
target values and the achieved RMSE.

\begin{table}[H]
	\centering
	\caption{Calibration test results for DEM parameters.}
		\label{tab:calibration_results}
	\begin{tabular}{lccc}
		\toprule
		Test Type          & Target Value         & Achieved Value       & RMSE
		\\
		\midrule
		Single-particle drop & Rebound height (mm)  & 10.2                  &
		9.8       \\
		Dry contact         & Restitution $e$      & 0.4                   &
		0.38      \\
		Pure-fluid sloshing & Frequency (Hz)       & 0.50                  &
		0.045     \\
		\bottomrule
	\end{tabular}
\end{table}

\section{Code Snippets}

\subsection{SPH Kernel Implementation}

The cubic spline kernel used in the SPH simulations is implemented as follows:

\begin{verbatim}
double W(double q, double h) {
	if (q < 1.0) {
		return (1.0 - 1.5 * q * q + 0.75 * q * q * q) / (7.0 * M_PI);
	} else if (q < 2.0) {
		return 0.25 * pow(2.0 - q, 3) / (7.0 * M_PI);
	} else {
		return 0.0;
	}
}
\end{verbatim}

\subsection{Coupling Algorithm Pseudocode}

The following pseudocode outlines the SPH--DEM coupling algorithm implemented in
the simulations:

\begin{verbatim}
for each time step:
	# Update SPH particles
	calculate_density();
	calculate_pressure();
	apply_viscous_forces();

	# Update DEM particles
	force_calculation();
	update_positions_velocities();

	# Coupling
	for each DEM particle:
		calculate_hydrodynamic_forces();
	apply_forces_to_SPH();
\end{verbatim}

\section*{Appendix Summary}
In summary, this appendix has provided supplementary material, including
additional data, code snippets, and detailed calculations, to support the
findings and methodologies presented in the thesis.

\end{document}